---
title: "lab_notebook_September"
output: rmarkdown::html_vignette rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{lab_notebook_september}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
```

# 2020 September 16

I flubbed my diarying in late August and early September because I was trying
to learn how to use the [www.tidyverse.org](tidyverse). Now I am reconstructing
the progress that I made on the analysis as I was using those tools.

## Powers Near 2 Need Large Degrees

True violet noise ($\alpha = 2$) requires a model with infinite degree. The
range of degrees that I used in my first round of simulations was good enough
for ($\alpha \lt 4/3$), but above that $\alpha$ the results seemed unstable. I
ran new simulations with VERY large degree ratios so that I could get more
insight into the behavior of models with large $\alpha$ values.

I need to find a way to summarize the simulations at each power so that I can
understand the effect of degree ratio at different powers. The original results
were fine for low degree ratios. However, as the power approaches 2 the degree
ratio must be very large, so I needed to rerun simulations with those high
degrees.

### Get the data from an SQLite database

The key variables in `high.degree` are `Slope.Error`, which is $|m - \alpha|$,
`Inverse.Sqrt.Size`, which is $1/\sqrt{N}$, and `Inverse.Degree.Ratio`, which
is $\sqrt{N}/K$. The model therefore has three parameters. A crude single
function for extracting the parameters and confidence intervals is:
```{r}
fo <- Slope.Error ~ Inverse.Sqrt.Size + Inverse.Degree.Ratio
fam <- Gamma(link="log")
column.names <- paste(rep(c("Intercept",
                            "Inv.Sqrt(Size)",
                            "Inv.Degree.Ratio"),
                      each=3),
                      c("lo", "B", "hi"),
                      sep=".")

extract.coefs.and.confints <- function(X) {
    mo <- glm(fo, fam, X)
    x <- as.numeric(rbind(coef(mo),
                       t(confint(mo)))[c(2, 1, 3), ])
    names(x) <- column.names
	data.frame(t(x))
}
```

I can then use some of the tidyverse's vaunted cleverness to fit models to each
power and extract the parameters and confidence intervals.

```{r}
if("power_law_model_info.Rdata" %in% dir("data-raw/")) {
    load("data-raw/power_model_info.Rdata")
}
else {
    con <- DBI::dbConnect(RSQLite::SQLite(),
                          "data-raw/power_law_results.sqlite3")
    high.degree <- tbl(con, "HighDegree")

    power.law.model.info <- high.degree %>%
                            group_by(Power) %>%
                            collect() %>%
                            group_modify(extract.coefs.and.confints)
	save(power.law.model.info, file="data-raw/power_model_info.Rdata")
```

Finally, plot the relationship between degree ratio, model power, and output
slope.
```{r fig.dim=c(6, 6), out.width="95%"}

plot(Slope ~ Power, contexts, las=1, cex=2, pch=16, col=size.colors[factor(Size)], xlab=expression(alpha), ylab="Degree Ratio Term")
grid()
lines(alphas, predict(loess(Slope ~ Power, contexts, span=1/3, weights=log(Size,2)), alphas), lwd=2)
legend("topleft", legend=sizes, bty="n", cex=4/3, pch=16, col=size.colors, title="Size", ncol=4)

```
The problem with this approach is that, when I try to save the `contexts`
tibble, EACH model gets saved with __its own copy of the full dataset__. Yikes!

I have to find a way to access the data without bringing R to its knees. Still
working on it ...

### Find Confidence Intervals for Parameters

```{r}
find.conf.int <- function(mo, ...) {
    tmp <- confint(mo, ...)
    interval.names <- paste(rep(c("lo", "hi"),
                                each=nrow(tmp)),
                            rownames(tmp),
                            sep=".")
    matrix(as.numeric(tmp),
           nrow=1,
           ncol=length(tmp),
           dimnames=list(NULL, interval.names))
}
#> pink.info <- data.frame(Power=pink.alphas,
                           t(sapply(pink.models, coef)),
                           AIC=sapply(pink.models, AIC))
#> pink.intervals <- data.frame(t(sapply(pink.models, find.conf.int)))
#> pink.info <- cbind(pink.info, pink.intervals)
```
