---
title: "tidy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Process the results using Tidyverse methods.

The following code uses a [www.tidyverse.org](tidyverse) approach to managing
the data.

First, open a database connection and read in the results as a `tibble`.
```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), "data-raw/power_law_results.sqlite3")
results <- tbl(con, "Centered")
contexts <- tibble(expand.grid(Power=alphas, Size=sizes))
```

Next, compute a generalized least-squares model with a Gamma-distributed error
for each collection of simulations with a common power.
```{r}
contexts$Model <- lapply(seq_len(nrow(contexts)), function(i){
    alpha <- contexts$Power[i]
    N <- contexts$Size[i]
    return(glm(Slope.Error ~ I(1/Degree.Ratio),
		   results %>% filter(abs(Power - alpha) < 0.02 & Size==N),
		   family=Gamma(link="log")))
}
contexts <- bind_cols(contexts,
                      contexts %>%
					  select(Model) %>%
					  rowwise() %>%
					  transmute(B=list(coef(Model))) %>%
					  transmute(Intercept=B[1], Slope=B[2])
					  )
```

Finally, plot the relationship between degree ratio, model power, and output
slope.
```{r fig.dim=c(6, 6), out.width="95%"}

plot(Slope ~ Power, contexts, las=1, cex=2, pch=16, col=size.colors[factor(Size)], xlab=expression(alpha), ylab="Degree Ratio Term")
grid()
lines(alphas, predict(loess(Slope ~ Power, contexts, span=1/3, weights=log(Size,2)), alphas), lwd=2)
legend("topleft", legend=sizes, bty="n", cex=4/3, pch=16, col=size.colors, title="Size", ncol=4)

```
